---
title: "MA615 Assignment 2 Part 1"
author: "Matthew Ciaramitaro"
date: "January 25, 2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Matrix problems

### 1. Suppose

$$ A = \begin{bmatrix}
1 & 1 & 3\\ 
5 & 2 & 6\\ 
-2 & -1 & -3
\end{bmatrix}$$
(a) Check that $A^3 = \mathbf{0}$  
```{r}
A <- matrix(c(1,1,3,5,2,6,-2,-1,-3), nrow = 3, byrow = T)
O <- matrix(rep(0,9), nrow=3, byrow=T)
sum(A %*% A %*% A == O) / length(A) == 1 #Check to see if A to the third is identical to the zero matrix
```
(b) Replace the third column of A by the 
sum of the second and third columns  

******   
First, produce A

```{r}
A <- matrix(c(1,1,3,5,2,6,-2,-1,-3), nrow = 3, byrow = TRUE)
A
```

Then, add the columns 2 and 3 and assign the sum to the third column


```{r}
A[,3] <- A[,2] + A[,3]

A
```


******

### 2. Create the following matrix B with 15 rows

$$B = \begin{bmatrix}
10 & -10 & 10\\
10 &  -10& 10\\
 \cdots & \cdots  & \cdots \\
10 & -10 & 10
\end{bmatrix} $$

Calculate the 3x3 matrix $B^TB$.  You can make this calculation with the function crossprod().  See the documentaion. 

```{r}
#first create B
B <- matrix(rep(c(10,-10, 10), length.out = 15), nrow=5, ncol=3, byrow = T)
B

```
```{r}
#now calculate B.T@B
out <- crossprod(B,B)
out
```
\newpage  


### 3. Create a 6 x 6 matrix matE  with every element equal to 0.  check what the functions row() and col() return when applied to matE.
Now, create the 6 x 6 matix:

$$\begin{bmatrix}
0 &1  & 0 & 0 & 0 & 0\\
1 &0 & 1 &0  & 0 &0 \\
0 & 1 & 0 & 1 & 0 & 0\\
0 & 0 & 1 & 0 & 1 & 0\\
0& 0 & 0 & 1 & 0 & 1\\
0& 0 & 0 & 0 & 1 & 0
\end{bmatrix} $$  

******
Here is matE, a 6x6 matrix of 0's followed by row(matE) and col(matE)
```{r}
matE <- matrix(rep(0,36), nrow = 6, byrow = TRUE)  

# Note what the functions row() and col() do
row(matE)
col(matE)

```


```{r}
# With a little experimentation you would see
# that the specified pattern is in the |1|'s
row(matE)-col(matE)
```
\newpage
```{r}
# so you use the locations of the 1's to modify matE 
matE[abs(row(matE)-col(matE))==1] <- 1
matE
```

******


### 4.  Look at the help for the function outer().  Now, create the following patterned matrix:

$$\begin{bmatrix}
0 & 1 & 2 & 3 & 4\\
1 & 2 & 3 & 4 & 5\\
2 & 3 & 4 & 5 & 6\\
3 & 4 & 5 & 6 & 7\\
4 & 5 & 6 & 7 & 8
\end{bmatrix} $$

```{r}

a <- 0:4
A <- outer(a,a,"+")
A
```

Use outer() a little more to make sure you get it.

```{r}
B <- outer(a,a, "*")
B

# and
b <- 5:10
C <- outer(a,b,"+")
C

# and finally  --  make sure you check the values.
D <- outer(b,a, "%%")
D
```


### 5.  Create the following patterned matrices.  Your solutions should be generalizable to enable creating larger matrices with the same structure.

(a)

$$\begin{bmatrix}
0 & 1 & 2 & 3 & 4\\
1 &2 & 3 & 4 & 0\\
2 & 3 & 4 & 0 & 1\\
3 & 4 & 0 & 1 & 2\\
4 & 0 & 1 & 2 & 3
\end{bmatrix}  $$
```{r}
a <- 0:4
A <- outer(a,a, "+")
A <- A %% length(a) #take sum modularly to generate the effect of increasing by 1 going down columns
A
```
(b)

$$ \begin{bmatrix}
0 & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9\\
1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 0\\
\vdots  & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
8 & 9 & 0 & 1 & 2 & 3 & 4 & 5 & 6 & 7\\
9 & 0 & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8
\end{bmatrix} $$
```{r}
#this is the same thing essentially with a different range
a  <-  0:9
A <- outer(a,a, "+")
A <- A %% length(a)
A
```

(c)

$$\begin{bmatrix}
0 & 8 & 7 & 6 & 5 & 4 &3  & 2 & 1\\
1 & 0 & 8 & 7 & 6 & 5 & 4 & 3 & 2\\
2 & 1 & 0 & 8 & 7 & 6 & 5 & 4 & 3\\
3 & 2 & 1 & 0 & 8 & 7 & 6 & 5 & 4\\
4 & 3 & 2 & 1 & 0 & 8 & 7 & 6 & 5\\
5 & 4 & 3 & 2 & 1 & 0 & 8 & 7 & 6\\
6 & 5 & 4 & 3 & 2 & 1 & 0 & 8 & 7\\
7 & 6 & 5 & 4 & 3 & 2 & 1 & 0 & 8\\
8 & 7 & 6 & 5 & 4 & 3 & 2 & 1 & 0
\end{bmatrix} $$


```{r}
#this is the same thing essentially with a different range
a  <-  0:8
A <- outer(a,a, "+")
A <- A %% length(a)
#Now the matrix here is essentially the correct matrix, but columns 2 to 9 are in the wrong order
A[,2:9] <- A[,9:2]
A
```




### 6.  Solve the following system of linear equations by setting up and solving the matrix equation Ax = y.

$x_1 + 2x_2 + 3x_3 + 4x_4 + 5x_5 =  7$  
$2x_1 +  x_2 + 2x_3 + 3x_4 + 4x_5 = -1$  
$3x_1 + 2x_2 +  x_3 + 2x_4 + 3x_5 = -3$  
$4x_1 + 3x_2 + 2x_3 +  x_4 + 2x_5 =  5$  
$5x_1 + 4x_2 + 3x_3 + 2x_4 +  x_5 = 17$  

```{r}
#first create the matrix A
A <-  matrix(rep(1:5, length.out=25), nrow=5, byrow = T)
A <-  abs(row(A) - col(A)) + 1
A
#x is the matrix we want to solve for, lets create y
y <- matrix(c(7,-1,-3,5,17), nrow=5, byrow = T)
y
#now solve
x <- solve(A,y)
x #this is the answer
#now check answer 
abs(A %*% x - y) < 1e-6 
```
 

### 7. Create a 6 x 10 matrix of random integers chosen from 1,2,...,10 by executing the following two lines of code:

```{r}
set.seed(75)  
aMat <- matrix(sample(10, size=60, replace=TRUE), nr=6)  
```

Use the matrix you have created to answer these questions:  

(a) Find the number of entries in each row which are greater than 4.
```{r}
sum(aMat > 4) #we expect this to be around 6/10 * 60 = 36 entries
```


(b) Which rows contain exactly two occurrences of the number seven?
```{r}
#answer is row 5
hist(which((aMat == 7), arr.ind = T)[,1]) #histogram illustrating number of 7s in each row
which(rowSums((aMat == 7)) == 2 ) #sums the rows of the logical vector of locations aMat == 7, gives rows which contain exactly 2 7s.

```
(c) Find those pairs of columns whose total (over both columns) is greater than 75.  The answer should be a matrix with two columns; so, for example, the row (1,2) in the output matrix means that the sum of columns 1 and 2 in the original matrix is greater than 75.  Repeating a column is permitted; so, for example, the final output matrix could contain the rows (1,2), (2,1), and (2,2).  

```{r}
allcolsums <- outer(colSums(aMat), colSums(aMat), "+")
allcolsums
sol <- which(allcolsums > 75, arr.ind = T)
sol
```


What if repetitions are not permitted? Then only (1,2) from (1,2),(2,1) and (2,2) would be permitted.
```{r}
allcolsums <- outer(colSums(aMat), colSums(aMat), "+")
allcolsums
sol <- allcolsums > 75
sol[lower.tri(sol)] <- F #remove below diagonal
diag(sol) <- F #remove the diagonal
sol <- which(sol, arr.ind = T)
sol

```

### 8. Calculate

(a) $\sum\limits_{i=1}^{20} \sum\limits_{j=1}^5 \frac{i^4}{(3+j)}$

******
```{r}
sum((1:20)^4) * sum(1/(3+(1:5)))

# or

sum(outer((1:20)^4, (3+(1:5)), "/"))

```
******

(b) $\sum\limits_{i=1}^{20} \sum\limits_{j=1}^5 \frac{i^4}{(3+ij)}$
```{r}
i <-1:20
j <- 1:5
sum(i^4/(3+i*j))


```



(c) $\sum\limits_{i=1}^{10} \sum\limits_{j=1}^i \frac{i^4}{(3+ij)}$
```{r}
i <-1:10
ij  <-  outer(i,i,"*") + 3
ij[upper.tri(ij)] <- Inf #because it wont count in the sum of the next step
sum(i^4/(ij))



```